---
title: "Reproducible Research - Peer Assessment 1"
author: "Doss Balaraman"
date: "Sunday, November 15, 2015"
output: html_document
---
```{r}
echo = TRUE  # Code visible
options(scipen = 1)  # Turning off scientific notation
data <- read.csv("activity.csv", colClasses = c("integer", "Date", "factor"))
data$month <- as.numeric(format(data$date, "%m"))
noNA <- na.omit(data)
rownames(noNA) <- 1:nrow(noNA)
head(noNA)
```
## Histogram of Total Number of Steps Taken Per Day:
```{r}
library(ggplot2)
ggplot(noNA, aes(date, steps)) + geom_bar(stat = "identity", colour = "steelblue", fill = "steelblue", width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day", x = "Date", y = "Total number of steps")
```

## Mean Total Number of Steps per Day
```{r}
totalSteps <- aggregate(noNA$steps, list(Date = noNA$date), FUN = "sum")$x
mean(totalSteps)
```

## Median total number of steps taken per day:
```{r}
median(totalSteps)
```

## Time Series Plot:
```{r}
avgSteps <- aggregate(noNA$steps, list(interval = as.numeric(as.character(noNA$interval))), FUN = "mean")
names(avgSteps)[2] <- "meanOfSteps"
ggplot(avgSteps, aes(interval, meanOfSteps)) + geom_line(color = "steelblue", size = 0.8) + labs(title = "Time Series Plot of the 5-minute Interval", x = "5-minute intervals", y = "Average Number of Steps Taken")
```

## Maximum Number of Steps among 5 min intervals:
```{r}
avgSteps[avgSteps$meanOfSteps == max(avgSteps$meanOfSteps), ]
```

## Number of rows with missing data:
```{r}
sum(is.na(data))
```

## Use 5 min avg to replace missing values:
```{r}
newData <- data 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanOfSteps
    }
}
head(newData)
sum(is.na(newData))
```

## Histogram with mean/median for new data after replacing missing values:
```{r}
ggplot(newData, aes(date, steps)) + geom_bar(stat = "identity",
                                             colour = "steelblue",
                                             fill = "steelblue",
                                             width = 0.7) + facet_grid(. ~ month, scales = "free") + labs(title = "Histogram of Total Number of Steps Taken Each Day (no missing data)", x = "Date", y = "Total number of steps")
```

## New Mean:
```{r}
newTotalSteps <- aggregate(newData$steps, 
                           list(Date = newData$date), 
                           FUN = "sum")$x
newMean <- mean(newTotalSteps)
newMean 
```

## New Median:
```{r}
newMedian <- median(newTotalSteps)
newMedian
```

## Comparison BEFORE and AFTER replacing missing values:
```{r}
oldMean <- mean(totalSteps)
oldMedian <- median(totalSteps)
newMean - oldMean
newMedian - oldMedian
```

## Weekdays and Weekends:
```{r}
newData$weekdays <- factor(format(newData$date, "%A"))
levels(newData$weekdays) 
levels(newData$weekdays) <- list(weekday = c("Monday", "Tuesday",
                                             "Wednesday", 
                                             "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newData$weekdays)
table(newData$weekdays)
```

## Time Series Plot of all weekdays and weekends:
```{r}
avgSteps <- aggregate(newData$steps, 
                      list(interval = as.numeric(as.character(newData$interval)), 
                           weekdays = newData$weekdays),
                      FUN = "mean")
names(avgSteps)[3] <- "meanOfSteps"
library(lattice)
xyplot(avgSteps$meanOfSteps ~ avgSteps$interval | avgSteps$weekdays, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Number of steps")
```
