---
title: "Reproducible Research - Peer Assessment 2 - Storm Data Research"
author: "Doss Balaraman"
date: "Sunday, November 22, 2015"
output: html_document
---
# Synonpsis

#### Analyze the impact of different weather events on public health and economy based on the storm 
#### database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. 
#### Use estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. Found that excessive heat and tornado are most harmful with respect 
#### to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

## Settings

```{r}
echo = TRUE  
options(scipen = 1) 
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
```
## Read Data & Collect meaningful data

```{r}
stormData<-read.csv("repdata-data-StormData.csv", sep = ",")

dim(stormData)
head(stormData, n = 2)
if (dim(stormData)[2] == 37) {
    stormData$year <- as.numeric(format(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(stormData$year, breaks = 30)
storm <- stormData[stormData$year >= 1995, ]
```

## Impact on Public Health

```{r}

sortHelper <- function(fieldName, top = 15, dataset = stormData) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortHelper("FATALITIES", dataset = storm)
injuries <- sortHelper("INJURIES", dataset = storm)
```

## Impact on Economy

#### Convert the property damage and crop damage data into comparable numerical forms according to the meaning of units described in the code book (Storm Events). Both PROPDMGEXP and CROPDMGEXP columns record a multiplier for each observation where we have Hundred (H), Thousand (K), Million (M) and Billion (B).

```{r}

convertHelper <- function(dataset = storm, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}

storm <- convertHelper(storm, "PROPDMGEXP", "propertyDamage")
storm <- convertHelper(storm, "CROPDMGEXP", "cropDamage")
names(storm)
options(scipen=999)
property <- sortHelper("propertyDamage", dataset = storm)
crop <- sortHelper("cropDamage", dataset = storm)

```

## Results

#### Two sorted lists of severe weather events below by the number of people badly affected.
#### fatalities
#### injuries
#### Graphs of total fatalities and total injuries affected by these severe weather events.

```{r}
fatalitiesPlot <- qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
injuriesPlot <- qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather\n Events in the U.S.\n from 1995 - 2011")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)

```
####
#### Graphs show excessive heat and tornado cause most fatalities;
#### tornato causes most injuries in the United States from 1995 to 2011.
#### 
#### Two lists to show impact by the amount of money cost by damages.
#### property
#### crop
#### Graphs to show total property damage and total crop damage affected by these severe weather events.

```{r}
propertyPlot <- qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage in US dollars")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events in\n the U.S. from 1995 - 2011")

cropPlot<- qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage in US dollars") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events in\n the U.S. from 1995 - 2011")

grid.arrange(propertyPlot, cropPlot, ncol = 2)

```
## RESULTS & Conclusion

####Based on the grpahs in the above analysis, flood and hurricane/typhoon cause most property damage; drought and flood causes most crop damage in the United States from 1995 to 2011.
#### Excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have more greater effect.